### 2022-04-08 chapters/fairness/justicia/framework.tex
enumeration approach complete

@@

\subsubsection{Measuring Fairness Metrics.}
As we compute the probability $\Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}]$ by solving the SSAT formula $ \Phi_\mathbf{a} $, we  use $ \Pr[\Phi_\mathbf{a}] $ to measure different fairness metrics. 
For that, we compute $ \Pr[\Phi_\mathbf{a}] $ for all compound groups $\mathbf{a} \in A$ that requires solving exponential (with $n$) number of SSAT instances. 
We elaborate this enumeration approach, namely {\justiciaenum}, in Algorithm~\ref{fairness_justicia_algo:enum}  (Line~\ref{fairness_justicia_algo:justicia_enum_begin}--\ref{fairness_justicia_algo:justicia_enum_end}).

\begin{algorithm}[t!]
	\caption{\justicia: SSAT-based Fairness Verifier}
	\label{fairness_justicia_algo:enum}
	\footnotesize
	\begin{algorithmic}[1]
		\Function{{\justiciaenum}}{$ X,A,\hat{Y} $}
		\label{fairness_justicia_algo:justicia_enum_begin}
		\State $ \phi_{\hat{Y}} := \mathsf{CNF}(\hat{Y} = 1) $
		%\State $ p_{i} = \mathsf{CalculateProb}(X_i), \forall X_i \in X $
		\ForAll{$\mathbf{a} \in A$ }
		\State $ p_{i} \leftarrow \mathsf{CalculateProb}(X_i | \mathbf{a}), \forall X_i \in X $
		\State $ \phi := \phi_{\hat{Y}} \wedge (\sensitive =\mathbf{a}) $
		\State $  \Phi_\mathbf{a} := \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \exists A_1,\dots, \exists A_{\numsensitive},  \phi $
		\State $ \Pr[\Phi_\mathbf{a}]  \leftarrow \mathsf{SSAT}(\Phi_\mathbf{a}) $ \Comment{returns a probability}
		\EndFor
		\State \Return $ \max_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}], \min_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] $
		\label{fairness_justicia_algo:justicia_enum_end}
		\EndFunction

		\Function{{\justicialearn}}{$ X,A,\hat{Y} $}
		\label{fairness_justicia_algo:justicia_learn_begin}
		\State $ \phi_{\hat{Y}} := \mathsf{CNF}(\hat{Y}  = 1) $
		\State $ p_{i} \leftarrow \mathsf{CalculateProb}(X_i), \forall X_i \in X $
		\State $  \Phi_\mathbf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \phi_{\hat{Y}} $
		\State $  \Phi'_\mathbf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \neg \phi_{\hat{Y}} $
		\State \Return $ \mathsf{SSAT}(\Phi_\mathbf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathbf{ER}) $
		\label{fairness_justicia_algo:justicia_learn_end}
		\EndFunction
	\end{algorithmic}

We calculate the ratio of the minimum and the maximum probabilities according to the definition of disparate impact. 
We compute statistical parity by taking the difference between the maximum and the minimum probabilities of all $ \Pr[\Phi_{\mathbf{a}}] $.
Moreover, to measure equalized odds, we compute two SSAT instances for each compound group with modified values of $ p_i $. 
Specifically, to compute TPR, we use the conditional probability $ p_i = \Pr[X_i|Y=1] $ on samples with class label $ Y = 1 $ and take the difference between the maximum and the minimum probabilities of all compound groups. In addition, to compute FPR, we use the conditional probability $ p_i = \Pr[X_i|Y=0] $ on samples with $ Y = 0 $ and take the difference similarly.
Thus, {\justiciaenum} allows us to compute different fairness metrics using a unified algorithmic framework.

\subsection{\red{Learning Fairness with ER-SSAT Encoding}}
\label{fairness_justicia_sec:learn_ssat}
In most practical problems, there can be exponentially many compound groups based on the different combinations of valuation to the sensitive features. 
Therefore, the enumeration approach may suffer from scalability issues. 
Hence, we propose efficient SSAT encodings to \textit{learn} the most favored group and the least favored group for given  $\alg$ and $ \mathcal{D} $, and to compute their \red{probability of positive prediction}s to measure different fairness metrics.

>>>>

\subsubsection{Measuring Fairness Metrics.}
As we compute $ \Pr[\Phi_\mathbf{a}] = \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $ by solving the SSAT formula $ \Phi_\mathbf{a} $, we  use $ \Pr[\Phi_\mathbf{a}] $ to measure different fairness metrics. To this end, we compute $ \Pr[\Phi_\mathbf{a}] $ for all compound groups $\mathbf{a} \in A$ by solving an exponential number (with $ \numsensitive $) of SSAT formulas. We elaborate this enumeration approach, namely {\justiciaenum}, in Algorithm~\ref{fairness_justicia_algo:enum}  (Line~\ref{fairness_justicia_algo:justicia_enum_begin}--\ref{fairness_justicia_algo:justicia_enum_end}).

\begin{algorithm}[t!]
	\caption{\justicia: SSAT-based Fairness Verifier}
	\label{fairness_justicia_algo:enum}
	\footnotesize
	\begin{algorithmic}[1]
		\Function{{\justiciaenum}}{$ \nonsensitive, \sensitive,\hat{Y} $}
		\label{fairness_justicia_algo:justicia_enum_begin}
		\State $ \phi_{\hat{Y}} := \mathsf{CNF}(\hat{Y} = 1) $
		%\State $ p_{i} = \mathsf{CalculateProb}(X_i), \forall X_i \in X $
		\ForAll{$\mathbf{a} \in \sensitive$ }
		\State $ p_{i} \leftarrow \Pr[X_i = 1|\sensitive = \mathbf{a}], \forall X_i \in \nonsensitive $
		\State $ \phi := \phi_{\hat{Y}} \wedge (\sensitive =\mathbf{a}) $
		\State $  \Phi_\mathbf{a} := \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \exists A_1,\dots, \exists A_{\numsensitive},  \phi $
		\State $ \Pr[\Phi_\mathbf{a}]  \leftarrow \mathsf{SSAT}(\Phi_\mathbf{a}) $ \Comment{returns a probability}
		\EndFor
		\State \Return $ \max_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}], \min_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $
		\label{fairness_justicia_algo:justicia_enum_end}
		\EndFunction

		\Function{{\justicialearn}}{$ \nonsensitive, \sensitive,\hat{Y} $}
		\label{fairness_justicia_algo:justicia_learn_begin}
		\State $ \phi_{\hat{Y}} := \mathsf{CNF}(\hat{Y}  = 1) $
		\State $ p_{i} \leftarrow \Pr[X_i=1], \forall X_i \in \nonsensitive $
		\State $  \Phi_\mathbf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \phi_{\hat{Y}} $
		\State $  \Phi'_\mathbf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \neg \phi_{\hat{Y}} $
		\State \Return $ \mathsf{SSAT}(\Phi_\mathbf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathbf{ER}) $
		\label{fairness_justicia_algo:justicia_learn_end}
		\EndFunction
	\end{algorithmic}

To measure the disparate impact of a classifier, we calculate the ratio between the minimum and the maximum conditional probability of positive prediction of the classifier, which are  $ \min_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $ and $ \max_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $, respectively. We compute statistical parity by taking the difference between $ \max_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $ and $ \min_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $.

Moreover, to compute equalized odds, we call {\justicia} twice, one for the distribution $ \mathcal{D} $ conditioned on $ Y = 1  $ and another for $ Y = 0 $. In both calls, we compute $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, Y = y ]  - \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, Y = y ] $ for $ y \in \{0,1\} $ and take the  maximum difference as the value of equalized odds. 
For measuring path-specific causal fairness, we compute  $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, \mathbf{Z}] $ and  $ \min_{ \mathbf{a}} \Pr[\hat{Y} =1, \mathbf{Z}| \sensitive = \mathbf{a} , \mathbf{Z}] $ by conditioning the distribution $ \mathcal{D} $ by mediator features $ \mathbf{Z} $ and take their difference. Thus, {\justiciaenum} allows us to compute different group and causal fairness metrics using a unified algorithmic framework.

@@

\subsection{\red{Inference Approach with ER-SSAT Encoding}}
\label{fairness_justicia_sec:learn_ssat}
In most practical problems, there can be exponentially many compound groups based on the different combinations of valuation to the sensitive features. 
Therefore, the enumeration approach may suffer from scalability issues. 
Hence, we propose efficient SSAT encodings to \textit{learn} the most favored group and the least favored group for given  $\alg$ and $ \mathcal{D} $, and to compute their \red{probability of positive prediction}s to measure different fairness metrics.
